---
output: 
  html_document:
params: 
  genename: 

  experiment_name: 
  treatment_variable: 
  control:
  treatment: 
  separator_variable: 
  num_meaningful_pcs: 
  

  
  filepath_Z_normalized: 
  filepath_loading_scores: 
  filepath_pca_eigenvalues:
  filepath_design_meaningful: 


title: "Summary of `r params$genename` behavior in `r params$experiment_name`"

---


```{r setup,echo=FALSE,include=FALSE}

#ABOUT: extracted gene-specific PCA summary from jcoffman008genesummary.Rmd (took out requirement of running DESEQ)
## summarizes the behavior of a gene of interest using the pca output of an experiment
#USAGE: run using Rscript runjcoffman008GeneSummary.R "genename"
#output: geneSummary.html 
#note: this script will use the parameters specified in runjcoffman008GeneSummary.R over the default params in this document

library(tidyr)
library(gridExtra)
library(ggplot2)
library(reshape2)
library(stringr)
library(knitr)

######Saving useful output files as dataframes. Creates objects nonspecific to the gene

#z-score (normalization that assumes normal distribution) from PCA 
zscore<- read.delim(file= params$filepath_Z_normalized)
##storing PCA gene loadings from the run with all samples
pca<- read.delim(file= params$filepath_loading_scores)
all_eigenvalues <- read.delim(file=params$filepath_pca_eigenvalues)
#convert loadings by multiplying by the square root of the corresponding vector to get -1 to 1 distribution
for(i in 1:params$num_meaningful_pcs){
  pca[,i+1]<- pca[,i+1]*sqrt(all_eigenvalues$eigenvalue[i]) 
} 
all_design<- read.delim(file=params$filepath_design_meaningful)
all_design$replicate <- as.factor(all_design$replicate)
all_design[,params$separator_variable]<- as.factor(all_design[,params$separator_variable])

#getting PCA results in format for plot 

#error-checking: making sure user did not input more meaningful PCs than recognized by rpca
design_col_num <- match("PC1", names(all_design))-1 #gives the number of columns before PC1 in the file  
if((design_col_num+params$num_meaningful_pcs)>length(names(all_design))){
  stop(paste("\n\n\n ********* WHY THE CODE FAILED!!: The number of meaningful principal components specified by the user in num_meaningful_pcs:  \n\t\t" ,params$num_meaningful_pcs, "\nexceeds the number of meaningful principal components in the rpca output.*********\n\n\n",sep=""))
}
long_pca<- melt(pca[c(1:(params$num_meaningful_pcs + 1))])


##Separating the labels of file-based dataframes
zscore<- separate(data = zscore, col = X, into = c("GeneID", "GeneName"), sep = "\\_",extra="merge")
pca<- separate(data = pca, col = X, into = c("GeneID", "GeneName"), sep = "\\_",extra="merge")
long_pca<- separate(data = long_pca, col = X, into = c("GeneID", "GeneName"), sep = "\\_",extra="merge")


#these lines are what allow control to come before treatment (i.e. gene knockout) in plots 
all_design[,params$treatment_variable]<- factor(all_design[,params$treatment_variable],levels=c(params$control,params$treatment))
all_design$X<- factor(all_design$X,levels=all_design$X[order(all_design[,params$treatment_variable])])

genename<- params$genename
#controlled error message if the gene is not in the output files that the plots use
if(genename %in% long_pca$GeneName==FALSE){
  stop(paste("\n\n\n ********* WHY THE CODE FAILED!!: You entered the following gene name:  \n\t\t" ,genename, "\nThis is not a recognized gene name or it is insufficiently expressed in experiment. Please check capitalization or enter another gene.*********\n\n\n",sep=""))
}

#adding the zscore normalized counts of this gene from each of these samples to the design file
design_genespecific<- all_design
design_genespecific$target_gene_zscore<- as.numeric(t(subset(zscore,GeneName==genename))[c(3:ncol(zscore)),])

###Facet plots to provide meaning to Principal Components. Only has to generate the plots if this is the first time running the script  

facetplotpng<- paste(params$experiment_name,"PCAfacetplots.png",sep="")

png(filename=facetplotpng, width=600, height=400)

if(!file.exists(facetplotpng)){
install.packages("patchwork",repos = "http://cran.us.r-project.org")
library(patchwork)




#adding the facet plot of PC1 as the first in the facet plot figure
facetplotfigure <- ggplot(data=all_design, aes(x=X, y=all_design[,(design_col_num+1)],fill=all_design[,params$treatment_variable]))+geom_bar(stat="identity", position=position_dodge())+ scale_fill_manual(params$treatment_variable,values = c("black", "red")) +
  facet_grid(. ~ all_design[,params$separator_variable], scales='free')+
  geom_text(aes(label = replicate), vjust = "outward", size=3,color="blue")+
  labs(y="Sample Loading")+
  theme_bw()+
  theme(axis.title.x=element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank(), panel.border = element_blank(), panel.grid.major = element_blank(),
        panel.grid.minor = element_blank())+ 
  ggtitle(paste("PC1: ", round(all_eigenvalues[1,]$variance.percent,2), "% of variance ",sep=""))

if(params$num_meaningful_pcs>1){
  #adding the rest of meaningful PCs to the facet plot figure
for( PC in c(2:params$num_meaningful_pcs)){
facetplotfigure <- facetplotfigure+ ggplot(data=all_design, aes_string(x=all_design$X, y=all_design[,(design_col_num+ PC )],fill=all_design[,params$treatment_variable]))+geom_bar(stat="identity", position=position_dodge())+ scale_fill_manual(params$treatment_variable,values = c("black", "red")) +
  facet_grid(. ~ all_design[,params$separator_variable], scales='free')+
  geom_text(aes(label = replicate), vjust = "outward", size=3, color="blue")+
  theme_bw()+
  labs(y="Sample Loading")+
  theme(axis.title.x=element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank(), panel.border = element_blank(), panel.grid.major = element_blank(),
        panel.grid.minor = element_blank())+ 
  ggtitle(paste("PC",PC,": ", round(all_eigenvalues[PC,]$variance.percent,2), "% of variance ",sep=""))

}
}

  facetplotfigure

  }else{
  include_graphics(facetplotpng)
  }
  dev.off()

###Creating plot to show relative PC loadings
plot_geneInPCA <- ggplot(data=long_pca, aes(x=variable,y=value))+geom_violin()+
  geom_point(data=subset(long_pca,GeneName==genename), color= "green",aes(y=value), size=4)+ theme(legend.position = "none", aspect.ratio = 1.5,panel.background = element_blank())+
  geom_hline(yintercept=0,linetype="dashed")+
  labs(x="",y= "Gene Loading")+
  ggtitle(paste("Relative ",genename," loadings in PCA of all samples",sep=""))

plot_expression <- ggplot(data=design_genespecific, aes(x=X, y=target_gene_zscore,color=design_genespecific[,params$treatment_variable]))+geom_point(size=5)+scale_color_manual(params$treatment_variable,values = c("black","red")) + 
  facet_grid(. ~ design_genespecific[,params$separator_variable], scales='free',switch="x")+
  geom_text(aes(label = replicate), vjust = -.2, size=5,color="blue")+
  theme_bw()+
  labs(x="Sample", y= "zscore norm count")+
  theme(aspect.ratio=2.5,
        axis.title.x=element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank(), panel.border = element_blank(), panel.grid.major = element_blank(),
        panel.grid.minor = element_blank())+ 
  ggtitle(paste("Z-score normalized ",genename, " counts separated by ", params$separator_variable, sep=""))

```
# {.tabset}

## All Results

``` {r displayAllPlots, out.width=c('40%', '30%','30%'), fig.show='hold', warning=FALSE, echo=FALSE}
include_graphics(facetplotpng)
plot_geneInPCA
plot_expression
```


## PCA of `r params$experiment_name`
PCA was conducted on `r params$experiment_name` to identify the dominating patterns of variance. The plots below split all samples by `r params$separator_variable` to demonstrate their loadings along each principal component, informing the pattern each principle component is capturing in relation to design variables. 

``` {r displayfacetplots, out.width="150%", warning=FALSE, echo=FALSE}
include_graphics(facetplotpng)

```

## `r genename` in PCA 

<div class = "row">

<div class = "col-md-6">
```{r PCAgenespecific, echo=FALSE}

plot_geneInPCA

PCA_percentile_report <- c()
for(PC in c(3:(params$num_meaningful_pcs+2))){
  PCA_percentile_report[[length(PCA_percentile_report)+1]]<- paste((round(mean(pca[,PC] < subset(pca,GeneName==genename)[,PC] ) ,2)*100)," percentile for PC", (PC-2), separator="")
}

```
</div>

<div class = "col-md-4">
<br><br>In the violin plots, `r genename` being at the top of the distribution of PC loadings indicates strong agreement with the patterns determined by that principal component. A loading near 0 indicates that `r genename` expression does not vary in relation to that pattern, and a loading at the bottom of the distribution suggests that `r genename` expression follows the opposite expression pattern. Compared to the distribution of loadings for all genes examined, `r genename` is loaded in the `r PCA_percentile_report`
</div>
</div>


## `r genename` expression

The plot below demonstrates how expression of `r genename` varies throughout `r params$experiment_name` by providing the normalized transcript counts from each sample.  

```{r expression,echo=FALSE,warning=FALSE}

plot_expression

```
